Generating Correlated Data for Omics Simulation

### 2025-05-23 manuscript/paper.qmd
Move back copula stuff to Methods
Otherwise, we lack too much context to explain our own methods

To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice.
We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body datasets.

@@

@@

1. Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 V_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $V = Z^T Z / (n-1)$ is the covariance matrix of $Z$.
2. Solve $A w = B$ and set $W$ to be the diagonal matrix with $w$ along its diagonal.
3. Set $U$ to be the $p \times k$ matrix with columns $u_i$.
4. Set $D$ to be the diagonal matrix with $D_{ii} = \sqrt{V_{ii} - (UW^2U^T)_{ii}}$, which is the remaining variance.

Steps 1 and 2 give that $u_i \Sigma_{sim} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
Step 4 ensures that $e_j \Sigma_{sim} e_j^T = 1$ for $j = 1, \ldots, p$.

>>>>

To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice, as in @fig-compare-to-real.
We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body datasets.

@@

Here, we provide an introduction to the Gaussian copula approach before describing our three methods for choosing $\Sigma_{sim}$.

### Multivariate normal distribution

We first discuss the simplest case, where our dataset is multivariate normally distributed.
The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
This matrix is conceptually simple, since $\Sigma_{ij}$ gives the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\widehat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
For large $n$, $\widehat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
In particular, $\widehat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
This means that $N(\mu, \widehat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.

The difficulty then is that we only know $\widehat \Sigma_{ref}$ from our reference dataset, but we need to choose a $\Sigma_{sim}$ with which we can simulate data and the obvious choice of $\widehat \Sigma_{ref}$ is inadequate.
The most common choice is to assume $\Sigma_{sim}$ is a diagonal matrix.
This is the situation we want to avoid where the generated data is independent: $x_i$ and $x_j$ have zero covariance unless $i = j$.
However, this has some nice properties, such as being simple and fast to simulate (just generate univariate normal data for each variable).

We describe three alternative approaches in Methods of how to choose a $\Sigma_{sim}$ that will produce data similar to the input data.
All three of these rely on a specific form of $\Sigma$, namely that
$$ \Sigma_{sim} = D^2 + UW^2U^T $$
where $D$ is a $p \times p$ diagonal matrix, $U$ is $p \times k$ for some $k \ll p$ and $W$ is a diagonal $k \times k$ matrix.
This is a combination of an independent part (the diagonal matrix) and a low-rank part ($U W^2 U^T$ is rank at most $k$).

Generating data for $\Sigma_{sim}$ of this form is highly efficient.
We use two basic facts about the multivariate distribution.
First, if $A$ is a matrix and $u ~ N(0, \Sigma)$ then $Au ~ N(0, A \Sigma A^T)$.
Second, if $u ~ N(0, \Sigma_1)$ and $v ~ N(0, \Sigma_2)$, then $u + v ~ N(0, \Sigma_1 + \Sigma_2)$.
Therefore, given matrices $D, U, W$ and $u ~ N(0, I_k)$ and $v ~ N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
$$ D v + U W u ~ N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
as desired.

In contrast, when given an arbitrary $\Sigma$, to generate a random value in $N(0, \Sigma)$, one must compute a matrix $V$ such that $VV^T = \Sigma$.
Then, using $v ~ N(0,I)$ , we have $Vv ~ N(0, \Sigma)$ as desired.
However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma$ is large.

Lastly, we emphasize that multivariate normal distributions do not capture all, or even most, types of possible dependence.
Indeed, we see this even in the 2-dimensional case where it is well known that correlation describes only a linear relationship between two variables while in reality they may have much more complex relations.
In higher dimensions, the problem is only worse.
So any method based off multivariate normal distributions are making large assumptions about distribution.
However, it is necessary to make some assumption like this.
In the next section, though, we see that "normal" part is actually not a large obstacle.

### Gaussian copula

Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
The idea of this approach is that by applying a normalizing transform and later reversing the transformation, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
This operates first by transforming each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.

Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
We describe this in detail:

1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
2.  Apply normalizing transform to $X$ by setting $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
5.  Set $z' = UWu + D v$.
6.  Output the vector $x'$ where $x'_i = F_i^{-1}(\Phi(z'))$.

The generated data vector $z'$ has covariance matrix $\Sigma_{sim} = D^2 + U W^2 U^T$.
Moreover, we require that $\Sigma$ satisfies that $\left(\Sigma_{sim}\right)_{ii}$ is approximately 1 for each $i$.
That guarantees that the output $x'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
This method is computationally efficient, taking hardly any more time or memory than simulations without dependence.

@@

1. Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 \left(\widehat \Sigma_{ref}\right)_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $\widehat \Sigma_{ref} = Z Z^T / (n-1)$ is the covariance matrix of $Z$.
2. Solve $A w = B$ and set $W$ to be the diagonal matrix with $w$ along its diagonal.
3. Set $U$ to be the $p \times k$ matrix with columns $u_i$.
4. Set $D$ to be the diagonal matrix with $D_{ii}^2 = \left(\widehat \Sigma_{ref}\right)_{ii} - (UW^2U^T)_{ii}$, which is the remaining variance.

Steps 1 and 2 give that $u_i \Sigma_{sim} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
Step 4 ensures that $\left(\Sigma_{sim}\right)_{jj} = 1$ for $j = 1, \ldots, p$.

### 2025-05-23 manuscript/supplemental/supplemental.qmd
Move back copula stuff to Methods
Otherwise, we lack too much context to explain our own methods

### Benchmark
```{r}
#| label: fig-benchmark
#| fig-cap: "Comparison of (a) run time and (b) memory usage for our methods, SPsimSeq, and vine copula (implemented by the rvinecopulib R package). 100 samples were generated with a varying number of genes. GSE151564 dataset was randomly subsetted to have the required number of genes to be used as the reference data set. Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
#| fig-width: 8
#| fig-height: 4
library(tidyverse)
library(patchwork)
METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula")
METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
benchmark <- read_tsv("../../processed/benchmark/results.txt") |>
    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints

g1 <- ggplot(benchmark, aes(x=n_genes, y=time, color=method)) +
    geom_line(linewidth=2) +
    labs(x = "N genes", y = "time (seconds)") + 
    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
    scale_x_log10() +
    scale_y_log10()
g2 <- ggplot(benchmark, aes(x=n_genes, y=max_rss, color=method)) +
    geom_line(linewidth=2) +
    labs(x = "N genes", y = "memory (MB)") +
    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
    scale_x_log10() +
    scale_y_log10()
g1 + g2 + plot_annotation(tag_levels="a") + plot_layout(guides = "collect")
```

@@

We used the rvinecopulib package to fit and generate random data from a vine copula distribution.
This package is a R wrapper of the high-performance vinecopulib C++ package.
This package expects data normalized to the range of \[0, 1\].
Therefore, we chose not to use the real RNA-seq count data and instead generated correlated random number as input which follow a multivariate normal distribution with covariance matrix constructed with a randomly generated singular vector matrix and singular values 1, 2, ..., 100.
These values were then normalized to the interval \[0,1\] by the `pnorm` function.
A vine copula was fit using just the "guassian" family.
While rvinecopula supports multi-threaded operation, we ran using a single thread as with the other methods.

>>>>

@@

We used the rvinecopulib package to fit and generate random data from a vine copula distribution.
This package is a R wrapper of the high-performance vinecopulib C++ package.
This package expects data scaled to the range of \[0, 1\].
Therefore, we chose not to use the real RNA-seq count data and instead generated correlated random number as input which follow a multivariate normal distribution with covariance matrix constructed with a randomly generated singular vector matrix and singular values 1, 2, ..., 100.
These values were then scaled to the interval \[0,1\] by the `pnorm` function.
A vine copula was fit using just the "guassian" family.
While rvinecopula supports multi-threaded operation, we ran using a single thread as with the other methods.

### Running mvrnorm

To assess a simple multivariate normal distribution simulation option, we ran the `mvrnorm` function from the MASS R package.
We set means to be the mean expression of each gene and used the sample covariance matrix as the `Sigma` parameter.
